---
title: "week4_Introduction to tidyverse"
author: "Macarena López-Mayorga"
date: 2026-03-27
format: pdf
---
First, we need to indicate where is located the file

```{r}
#| eval: FALSE
FilePath <- file.path("C:/Users/Macarena/CytometryInR/course/week04/04_IntroToTidyverse/data/Dataset.csv")
FilePath
```

```{r}
#| eval: FALSE
library(utils)
```


```{r}
#| eval: FALSE
Data <- read.csv(file=FilePath, check.names=FALSE)
colnames(Data)
```
 [1] "bid"               "timepoint"         "Condition"         "Date"              "infant_sex"        "ptype"            
 [7] "root"              "singletsFSC"       "singletsSSC"       "singletsSSCB"      "CD45"              "NotMonocytes"     
[13] "nonDebris"         "lymphocytes"       "live"              "Dump+"             "Dump-"             "Tcells"           
[19] "Vd2+"              "Vd2-"              "Va7.2+"            "Va7.2-"            "CD4+"              "CD4-"             
[25] "CD8+"              "CD8-"              "Tcells_count"      "lymphocytes_count" "Monocytes"         "Debris"           
[31] "CD45_count"

when we put check.names into TRUE, then the "+" or "-" symbols dissapear

```{r}
#| eval: FALSE
Data <- read.csv(file=FilePath, check.names=TRUE)
colnames(Data)
```
 [1] "bid"               "timepoint"         "Condition"         "Date"              "infant_sex"        "ptype"            
 [7] "root"              "singletsFSC"       "singletsSSC"       "singletsSSCB"      "CD45"              "NotMonocytes"     
[13] "nonDebris"         "lymphocytes"       "live"              "Dump."             "Dump..1"           "Tcells"           
[19] "Vd2."              "Vd2..1"            "Va7.2."            "Va7.2..1"          "CD4."              "CD4..1"           
[25] "CD8."              "CD8..1"            "Tcells_count"      "lymphocytes_count" "Monocytes"         "Debris"           
[31] "CD45_count"   

To see some rows:
```{r}
#| eval: FALSE
head(Data, 3)
```
bid timepoint Condition       Date infant_sex  ptype    root singletsFSC singletsSSC singletsSSCB      CD45 NotMonocytes nonDebris
1 INF0052         0      Ctrl 2025-07-26       Male HEU-hi 2098368     1894070     1666179      1537396 0.5952943    0.8820349 0.8627649
2 INF0100         0      Ctrl 2025-07-26       Male HEU-lo 2020184     1791890     1697083      1579098 0.9106762    0.9052256 0.8602660
3 INF0100         4      Ctrl 2025-07-26       Male HEU-lo 1155040     1033320      875465       845446 0.9705765    0.9845400 0.9578793
  lymphocytes      live      Dump+     Dump-    Tcells        Vd2+      Vd2-     Va7.2+    Va7.2-      CD4+      CD4-      CD8+
1   0.6420138 0.9020581 0.21090996 0.6911482 0.2804264 0.008120361 0.9918796 0.01448070 0.9773989 0.6341164 0.3432825 0.2734826
2   0.2145848 0.8908981 0.06252775 0.8283703 0.6748298 0.007265620 0.9927344 0.01577499 0.9769594 0.6119112 0.3650482 0.3357696
3   0.7403110 0.8757665 0.20023803 0.6755285 0.6119129 0.004651313 0.9953487 0.01579402 0.9795547 0.6639621 0.3155925 0.2862104
        CD8- Tcells_count lymphocytes_count  Monocytes     Debris CD45_count
1 0.06979990       164771            587573 0.11796509 0.13723513     915203
2 0.02927858       208241            308583 0.09477437 0.13973396    1438047
3 0.02938209       371723            607477 0.01545999 0.04212072     820570

To visualize the table, we can click into the grid, in the rigth panel, or using the next code:
```{r}
#| eval: FALSE
View(Data)
```
To see the type of data:

```{r}
#| eval: FALSE
class(Data)
```
[1] "data.frame"

And see the variables:
```{r}
#| eval: FALSE
str(Data)
```
'data.frame':	196 obs. of  31 variables:
 $ bid              : chr  "INF0052" "INF0100" "INF0100" "INF0100" ...
 $ timepoint        : int  0 0 4 9 0 4 9 4 9 0 ...
 $ Condition        : chr  "Ctrl" "Ctrl" "Ctrl" "Ctrl" ...
 $ Date             : chr  "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" ...
 $ infant_sex       : chr  "Male" "Male" "Male" "Male" ...
 $ ptype            : chr  "HEU-hi" "HEU-lo" "HEU-lo" "HEU-lo" ...
 $ root             : int  2098368 2020184 1155040 358624 1362216 1044808 1434840 972056 1521928 2363512 ...
 $ singletsFSC      : int  1894070 1791890 1033320 328624 1206309 917398 1265022 875707 1359574 2136616 ...
 $ singletsSSC      : int  1666179 1697083 875465 289327 1032946 735579 988445 767323 1175755 1875394 ...
 $ singletsSSCB     : int  1537396 1579098 845446 276289 982736 685592 940454 718000 1097478 1732620 ...
 $ CD45             : num  0.595 0.911 0.971 0.982 0.957 ...
 $ NotMonocytes     : num  0.882 0.905 0.985 0.986 0.956 ...
 $ nonDebris        : num  0.863 0.86 0.958 0.941 0.841 ...
 $ lymphocytes      : num  0.642 0.215 0.74 0.651 0.705 ...
 $ live             : num  0.902 0.891 0.876 0.915 0.895 ...
 $ Dump+            : num  0.2109 0.0625 0.2002 0.2147 0.3383 ...
 $ Dump-            : num  0.691 0.828 0.676 0.701 0.557 ...
 $ Tcells           : num  0.28 0.675 0.612 0.631 0.44 ...
 $ Vd2+             : num  0.00812 0.00727 0.00465 0.01135 0.00475 ...
 $ Vd2-             : num  0.992 0.993 0.995 0.989 0.995 ...
 $ Va7.2+           : num  0.0145 0.0158 0.0158 0.017 0.0133 ...
 $ Va7.2-           : num  0.977 0.977 0.98 0.972 0.982 ...
 $ CD4+             : num  0.634 0.612 0.664 0.438 0.739 ...
 $ CD4-             : num  0.343 0.365 0.316 0.534 0.243 ...
 $ CD8+             : num  0.273 0.336 0.286 0.486 0.195 ...
 $ CD8-             : num  0.0698 0.0293 0.0294 0.0476 0.0476 ...
 $ Tcells_count     : int  164771 208241 371723 111552 291777 271870 487937 220634 415867 184930 ...
 $ lymphocytes_count: int  587573 308583 607477 176662 663667 510730 726238 451047 710964 652155 ...
 $ Monocytes        : num  0.118 0.0948 0.0155 0.0145 0.0444 ...
 $ Debris           : num  0.1372 0.1397 0.0421 0.0587 0.1592 ...
 $ CD45_count       : int  915203 1438047 820570 271304 940733 675857 921660 701657 1066884 1017713 ...

To see that using glimpse function (if we don't know from what package is, we can use the next code to figure out)
 ```{r}
 #| eval: FALSE
 ??glimpse
 ```

 It's from dplyr package, so we have to activate it
 
 ```{r}
 #| eval: FALSE
 library(dplyr)
glimpse(Data)
 ```
 Rows: 196
Columns: 31
$ bid               <chr> "INF0052", "INF0100", "INF0100", "INF0100", "INF0179", "INF0179", "INF0179", "INF0186", "INF0186", "INF0052", …
$ timepoint         <int> 0, 0, 4, 9, 0, 4, 9, 4, 9, 0, 0, 4, 9, 0, 4, 9, 4, 9, 0, 0, 4, 0, 4, 9, 4, 9, 0, 4, 9, 0, 9, 0, 4, 9, 0, 4, 9,…
$ Condition         <chr> "Ctrl", "Ctrl", "Ctrl", "Ctrl", "Ctrl", "Ctrl", "Ctrl", "Ctrl", "Ctrl", "PPD", "PPD", "PPD", "PPD", "PPD", "PP…
$ Date              <chr> "2025-07-26", "2025-07-26", "2025-07-26", "2025-07-26", "2025-07-26", "2025-07-26", "2025-07-26", "2025-07-26"…
$ infant_sex        <chr> "Male", "Male", "Male", "Male", "Male", "Male", "Male", "Female", "Female", "Male", "Male", "Male", "Male", "M…
$ ptype             <chr> "HEU-hi", "HEU-lo", "HEU-lo", "HEU-lo", "HU", "HU", "HU", "HEU-hi", "HEU-hi", "HEU-hi", "HEU-lo", "HEU-lo", "H…
$ root              <int> 2098368, 2020184, 1155040, 358624, 1362216, 1044808, 1434840, 972056, 1521928, 2363512, 2049112, 1063496, 7883…
$ singletsFSC       <int> 1894070, 1791890, 1033320, 328624, 1206309, 917398, 1265022, 875707, 1359574, 2136616, 1821676, 946587, 714198…
$ singletsSSC       <int> 1666179, 1697083, 875465, 289327, 1032946, 735579, 988445, 767323, 1175755, 1875394, 1717636, 796056, 626387, …
$ singletsSSCB      <int> 1537396, 1579098, 845446, 276289, 982736, 685592, 940454, 718000, 1097478, 1732620, 1597085, 767297, 600011, 1…
$ CD45              <dbl> 0.5952943, 0.9106762, 0.9705765, 0.9819573, 0.9572591, 0.9858006, 0.9800160, 0.9772382, 0.9721234, 0.5873838, …
$ NotMonocytes      <dbl> 0.8820349, 0.9052256, 0.9845400, 0.9855070, 0.9556272, 0.9622894, 0.9805210, 0.9490663, 0.9581698, 0.8619837, …
$ nonDebris         <dbl> 0.8627649, 0.8602660, 0.9578793, 0.9412615, 0.8407837, 0.9499243, 0.9657759, 0.9189704, 0.9186763, 0.8429685, …
$ lymphocytes       <dbl> 0.6420138, 0.2145848, 0.7403110, 0.6511588, 0.7054786, 0.7556776, 0.7879674, 0.6428312, 0.6663930, 0.6408044, …
$ live              <dbl> 0.9020581, 0.8908981, 0.8757665, 0.9153242, 0.8952140, 0.9673056, 0.9297420, 0.8756515, 0.8749529, 0.9009254, …
$ `Dump+`           <dbl> 0.21090996, 0.06252775, 0.20023803, 0.21469246, 0.33831877, 0.36947311, 0.23863114, 0.32516789, 0.22875701, 0.…
$ `Dump-`           <dbl> 0.6911482, 0.8283703, 0.6755285, 0.7006317, 0.5568953, 0.5978325, 0.6911109, 0.5504837, 0.6461959, 0.6934931, …
$ Tcells            <dbl> 0.2804264, 0.6748298, 0.6119129, 0.6314431, 0.4396437, 0.5323165, 0.6718693, 0.4891597, 0.5849340, 0.2835676, …
$ `Vd2+`            <dbl> 0.008120361, 0.007265620, 0.004651313, 0.011348967, 0.004753630, 0.005160555, 0.002631487, 0.009300470, 0.0179…
$ `Vd2-`            <dbl> 0.9918796, 0.9927344, 0.9953487, 0.9886510, 0.9952464, 0.9948394, 0.9973685, 0.9906995, 0.9820592, 0.9925918, …
$ `Va7.2+`          <dbl> 0.014480704, 0.015774991, 0.015794019, 0.017023451, 0.013321818, 0.012461838, 0.013563226, 0.038770090, 0.0367…
$ `Va7.2-`          <dbl> 0.9773989, 0.9769594, 0.9795547, 0.9716276, 0.9819246, 0.9823776, 0.9838053, 0.9519294, 0.9452878, 0.9775212, …
$ `CD4+`            <dbl> 0.6341164, 0.6119112, 0.6639621, 0.4378944, 0.7392563, 0.5876264, 0.6994141, 0.5724730, 0.5119209, 0.6340345, …
$ `CD4-`            <dbl> 0.3432825, 0.3650482, 0.3155925, 0.5337331, 0.2426682, 0.3947512, 0.2843912, 0.3794565, 0.4333669, 0.3434867, …
$ `CD8+`            <dbl> 0.2734826, 0.3357696, 0.2862104, 0.4861231, 0.1950634, 0.3571376, 0.2580784, 0.3555526, 0.4138799, 0.2744119, …
$ `CD8-`            <dbl> 0.06979990, 0.02927858, 0.02938209, 0.04761008, 0.04760485, 0.03761357, 0.02631282, 0.02390384, 0.01948700, 0.…
$ Tcells_count      <int> 164771, 208241, 371723, 111552, 291777, 271870, 487937, 220634, 415867, 184930, 211987, 326378, 238021, 294549…
$ lymphocytes_count <int> 587573, 308583, 607477, 176662, 663667, 510730, 726238, 451047, 710964, 652155, 314717, 544883, 366784, 663169…
$ Monocytes         <dbl> 0.11796509, 0.09477437, 0.01545999, 0.01449297, 0.04437285, 0.03771064, 0.01947898, 0.05093372, 0.04183023, 0.…
$ Debris            <dbl> 0.13723513, 0.13973396, 0.04212072, 0.05873854, 0.15921627, 0.05007568, 0.03422412, 0.08102962, 0.08132374, 0.…
$ CD45_count        <int> 915203, 1438047, 820570, 271304, 940733, 675857, 921660, 701657, 1066884, 1017713, 1447451, 745037, 589379

This function is similar to str() function but more tidy and with the dimensions of the table at the top. To see only the dimensions, we can also use de dim() function, which maintains the row followed by column position convention of base R (first row, second column)

```{r}
#| eval: FALSE
dim(Data)
```
[1] 196  31

To verify the column value type (character, for instance), we can use the functions is.character, is.numeric or is.logical:

```{r}
#| eval: FALSE
is.character(Data$bid)
```
[1] TRUE

```{r}
#| eval: FALSE
is.numeric(Data$bid)
```
[1] FALSE

```{r}
#| eval: FALSE
is.character(Data$Tcells_count)
```
[1] FALSE
For numeric columns using the is.numeric() function, we can also be subtype specific using either is.integer() or is.double():

```{r}
#| eval: FALSE
is.numeric(Data$Tcells_count)
```
[1] TRUE

```{r}
#| eval: FALSE
is.integer(Data$Tcells_count)
```
[1] TRUE


```{r}
#| eval: FALSE
is.double(Data$Tcells_count)
```
[1] FALSE

Like with keywords, special characters (as & or spaces) will need tick marks to run the function

```{r}
#| eval: FALSE
is.numeric(Data$`CD8-`)
```
[1] TRUE


```{r}
#| eval: FALSE
library(dplyr)
```
The select() function is used to “select” a column from a data.frame type object. The simplest way to use it is giving the name of the data.frame object in first place and then, the name of the column, surrounded by " ":

```{r}
#| eval: FALSE
DateColumn <- select(Data, "Date")
DateColumn[1:10,]
```
[1] "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26"

This results in the column being selected, resulting in the new object containing only that subsetted out column from the original Data object.

In general, the function select() is used with the implementation of pipe operators, not like we used previously:

```{r}
#| eval: FALSE
DateColumn <- Data |> select("Date")
DateColumn[1:10,]
```
 [1] "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26"

 The main advantage of the pipe operators is that we can link different functions. For example, the nrow() function determine the number of rows. We can link the function select() and nrow() with the pipe operators:
 
 ```{r}
 #| eval: FALSE
 Data |> select("Date") |> nrow()
 ```
 [1] 196

 CURIOUS NOTE: Despite we need "" for install.packages() function, for instance, in select() function, the function still run without the "" in date:

 ```{r}
 #| eval: FALSE
 Data |> select(Date) |> head(3)
 ```
         Date
1 2025-07-26
2 2025-07-26
3 2025-07-26

This is due to dplyr; It's now in charge and it's able to infer that Date is a column, with no need of "".
For select more than one column and position them anywhere within the data.frame:

```{r}
#| eval: FALSE
Subset <- Data |> select(bid, timepoint, Condition, Tcells, `CD8+`, `CD4+`)
head(Subset, 3)
```
      bid timepoint Condition    Tcells      CD8+      CD4+
1 INF0052         0      Ctrl 0.2804264 0.2734826 0.6341164
2 INF0100         0      Ctrl 0.6748298 0.3357696 0.6119112
3 INF0100         4      Ctrl 0.6119129 0.2862104 0.6639621

The order in which we selected the columns will dictate their position in the subsetted data.frame object:

```{r}
#| eval: FALSE
Subset <- Data |> select(bid, Tcells, `CD8+`, `CD4+`, timepoint, Condition)
head(Subset, 3)
```
      bid    Tcells      CD8+      CD4+ timepoint Condition
1 INF0052 0.2804264 0.2734826 0.6341164         0      Ctrl
2 INF0100 0.6748298 0.3357696 0.6119112         0      Ctrl
3 INF0100 0.6119129 0.2862104 0.6639621         4      Ctrl

The relocate() function is used when we want to move just one column. We can do this with select() function too, but it would be a waste of time. Let’s say we wanted to move the Tcells column from its current location to the second column position (right after the bid column). The line of code to do so would look like:

```{r}
#| eval: FALSE
Data |> relocate(Tcells, .after=bid) |> head(3)
```
      bid    Tcells timepoint Condition       Date infant_sex  ptype    root singletsFSC singletsSSC singletsSSCB      CD45 NotMonocytes
1 INF0052 0.2804264         0      Ctrl 2025-07-26       Male HEU-hi 2098368     1894070     1666179      1537396 0.5952943    0.8820349
2 INF0100 0.6748298         0      Ctrl 2025-07-26       Male HEU-lo 2020184     1791890     1697083      1579098 0.9106762    0.9052256
3 INF0100 0.6119129         4      Ctrl 2025-07-26       Male HEU-lo 1155040     1033320      875465       845446 0.9705765    0.9845400
  nonDebris lymphocytes      live      Dump+     Dump-        Vd2+      Vd2-     Va7.2+    Va7.2-      CD4+      CD4-      CD8+
1 0.8627649   0.6420138 0.9020581 0.21090996 0.6911482 0.008120361 0.9918796 0.01448070 0.9773989 0.6341164 0.3432825 0.2734826
2 0.8602660   0.2145848 0.8908981 0.06252775 0.8283703 0.007265620 0.9927344 0.01577499 0.9769594 0.6119112 0.3650482 0.3357696
3 0.9578793   0.7403110 0.8757665 0.20023803 0.6755285 0.004651313 0.9953487 0.01579402 0.9795547 0.6639621 0.3155925 0.2862104
        CD8- Tcells_count lymphocytes_count  Monocytes     Debris CD45_count
1 0.06979990       164771            587573 0.11796509 0.13723513     915203
2 0.02927858       208241            308583 0.09477437 0.13973396    1438047
3 0.02938209       371723            607477 0.01545999 0.04212072     820570

Similar to what we saw with select(), this approach can also be used for more than 1 column:

```{r}
#| eval: FALSE
Data |> relocate(Tcells, Monocytes, .after=bid) |> head(3)
```
      bid    Tcells  Monocytes timepoint Condition       Date infant_sex  ptype    root singletsFSC singletsSSC singletsSSCB      CD45
1 INF0052 0.2804264 0.11796509         0      Ctrl 2025-07-26       Male HEU-hi 2098368     1894070     1666179      1537396 0.5952943
2 INF0100 0.6748298 0.09477437         0      Ctrl 2025-07-26       Male HEU-lo 2020184     1791890     1697083      1579098 0.9106762
3 INF0100 0.6119129 0.01545999         4      Ctrl 2025-07-26       Male HEU-lo 1155040     1033320      875465       845446 0.9705765
  NotMonocytes nonDebris lymphocytes      live      Dump+     Dump-        Vd2+      Vd2-     Va7.2+    Va7.2-      CD4+      CD4-
1    0.8820349 0.8627649   0.6420138 0.9020581 0.21090996 0.6911482 0.008120361 0.9918796 0.01448070 0.9773989 0.6341164 0.3432825
2    0.9052256 0.8602660   0.2145848 0.8908981 0.06252775 0.8283703 0.007265620 0.9927344 0.01577499 0.9769594 0.6119112 0.3650482
3    0.9845400 0.9578793   0.7403110 0.8757665 0.20023803 0.6755285 0.004651313 0.9953487 0.01579402 0.9795547 0.6639621 0.3155925
       CD8+       CD8- Tcells_count lymphocytes_count     Debris CD45_count
1 0.2734826 0.06979990       164771            587573 0.13723513     915203
2 0.3357696 0.02927858       208241            308583 0.13973396    1438047
3 0.2862104 0.02938209       371723            607477 0.04212072     820570

To rename columns, we can proceed calling a subset of data and then, specifying the column (third, first, whatever). Once we confirm the column that we want to rename, next we can change it:

```{r}
#| eval: FALSE
colnames(Subset)
```
[1] "bid"       "Tcells"    "CD8+"      "CD4+"      "timepoint" "Condition"

Now, we specify the column in which we are interested:
```{r}
#| eval: FALSE
colnames(Subset)[3]
```
And we change the name of that particular column:

```{r}
#| eval: FALSE
colnames(Subset)[3] <- "CD8Positive"
colnames(Subset)
```
[1] "bid"         "Tcells"      "CD8Positive" "CD4+"        "timepoint"   "Condition"  
We can also do that in a different way:

```{r}
#| eval: FALSE
Renamed <- Subset |> rename(CD4_Positive = `CD4+`)
colnames(Renamed)
```
[1] "bid"          "Tcells"       "CD8Positive"  "CD4_Positive" "timepoint"    "Condition"   

If we want to rename several columns at once. we can do it using a comma between them:

```{r}
#| eval: FALSE
Renamed_Multiple <- Subset |> rename(specimen = bid, timepoint_months = timepoint, stimulation = Condition, CD4Positive=`CD4+`)
colnames(Renamed_Multiple)
```
[1] "specimen"         "Tcells"           "CD8Positive"      "CD4Positive"      "timepoint_months" "stimulation"   

To see data from a particular column, we use the function pull():

```{r}
#| eval: FALSE
Data |> pull(Date) |> head(5)
```
[1] "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26" "2025-07-26"

If we also want to see the unique values in a column, we can use the unique() function:

```{r}
#| eval: FALSE
Data |> pull(Date) |> unique()
```
[1] "2025-07-26" "2025-07-29" "2025-07-31" "2025-08-05" "2025-08-07" "2025-08-22" "2025-08-28" "2025-08-30"

The Condition column in this Dataset appears to be indicating whether the samples were stimulated. Let’s see how many unique values are contained within that column

```{r}
#| eval: FALSE
Data |> pull(Condition) |> unique() 
```
[1] "Ctrl" "PPD"  "SEB" 

If we want to subset just one treatment, we can use the function filter(): IMPORTANT NOTE: In this case, "" are necessary as we are asking for a matching character value.

```{r}
#| eval: FALSE
PPDOnly <- Data |> filter(Condition == "PPD")
head(PPDOnly, 5)
```
      bid timepoint Condition       Date infant_sex  ptype    root singletsFSC singletsSSC singletsSSCB      CD45 NotMonocytes nonDebris
1 INF0052         0       PPD 2025-07-26       Male HEU-hi 2363512     2136616     1875394      1732620 0.5873838    0.8619837 0.8429685
2 INF0100         0       PPD 2025-07-26       Male HEU-lo 2049112     1821676     1717636      1597085 0.9063081    0.9251961 0.8771889
3 INF0100         4       PPD 2025-07-26       Male HEU-lo 1063496      946587      796056       767297 0.9709891    0.9848719 0.9556049
4 INF0100         9       PPD 2025-07-26       Male HEU-lo  788368      714198      626387       600011 0.9822803    0.9842139 0.8123041
5 INF0179         0       PPD 2025-07-26       Male     HU 1380336     1242311     1047081      1000877 0.9470275    0.9575685 0.9134438
  lymphocytes      live      Dump+     Dump-    Tcells        Vd2+      Vd2-     Va7.2+    Va7.2-      CD4+      CD4-      CD8+
1   0.6408044 0.9009254 0.20743228 0.6934931 0.2835676 0.007408209 0.9925918 0.01507057 0.9775212 0.6340345 0.3434867 0.2744119
2   0.2174284 0.8929673 0.06181426 0.8311531 0.6735798 0.007137230 0.9928628 0.01671801 0.9761448 0.6145707 0.3615741 0.3312279
3   0.7313503 0.8782307 0.20727202 0.6709587 0.5989873 0.005254643 0.9947454 0.01609790 0.9786475 0.6559480 0.3226994 0.2912084
4   0.6223228 0.9566639 0.23164587 0.7250180 0.6489405 0.011935922 0.9880641 0.01855298 0.9695111 0.4306889 0.5388222 0.4908558
5   0.6996502 0.8856898 0.33186111 0.5538287 0.4441538 0.004382972 0.9956170 0.01297237 0.9826447 0.7499194 0.2327253 0.1850897
        CD8- Tcells_count lymphocytes_count  Monocytes     Debris CD45_count
1 0.06907479       184930            652155 0.13801632 0.15703150    1017713
2 0.03034620       211987            314717 0.07480391 0.12281107    1447451
3 0.03149109       326378            544883 0.01512811 0.04439511     745037
4 0.04796636       238021            366784 0.01578611 0.18769586     589379
5 0.04763554       294549            663169 0.04243146 0.08655621     947858

Although the code previously used works, it can generate error, specially with character values. For that reason, it is better to use  %in% operator to extract rows from a column.

```{r}
#| eval: FALSE
Data |> filter(Condition %in% "PPD") |> head(5)
```
      bid timepoint Condition       Date infant_sex  ptype    root singletsFSC singletsSSC singletsSSCB      CD45 NotMonocytes nonDebris
1 INF0052         0       PPD 2025-07-26       Male HEU-hi 2363512     2136616     1875394      1732620 0.5873838    0.8619837 0.8429685
2 INF0100         0       PPD 2025-07-26       Male HEU-lo 2049112     1821676     1717636      1597085 0.9063081    0.9251961 0.8771889
3 INF0100         4       PPD 2025-07-26       Male HEU-lo 1063496      946587      796056       767297 0.9709891    0.9848719 0.9556049
4 INF0100         9       PPD 2025-07-26       Male HEU-lo  788368      714198      626387       600011 0.9822803    0.9842139 0.8123041
5 INF0179         0       PPD 2025-07-26       Male     HU 1380336     1242311     1047081      1000877 0.9470275    0.9575685 0.9134438
  lymphocytes      live      Dump+     Dump-    Tcells        Vd2+      Vd2-     Va7.2+    Va7.2-      CD4+      CD4-      CD8+
1   0.6408044 0.9009254 0.20743228 0.6934931 0.2835676 0.007408209 0.9925918 0.01507057 0.9775212 0.6340345 0.3434867 0.2744119
2   0.2174284 0.8929673 0.06181426 0.8311531 0.6735798 0.007137230 0.9928628 0.01671801 0.9761448 0.6145707 0.3615741 0.3312279
3   0.7313503 0.8782307 0.20727202 0.6709587 0.5989873 0.005254643 0.9947454 0.01609790 0.9786475 0.6559480 0.3226994 0.2912084
4   0.6223228 0.9566639 0.23164587 0.7250180 0.6489405 0.011935922 0.9880641 0.01855298 0.9695111 0.4306889 0.5388222 0.4908558
5   0.6996502 0.8856898 0.33186111 0.5538287 0.4441538 0.004382972 0.9956170 0.01297237 0.9826447 0.7499194 0.2327253 0.1850897
        CD8- Tcells_count lymphocytes_count  Monocytes     Debris CD45_count
1 0.06907479       184930            652155 0.13801632 0.15703150    1017713
2 0.03034620       211987            314717 0.07480391 0.12281107    1447451
3 0.03149109       326378            544883 0.01512811 0.04439511     745037
4 0.04796636       238021            366784 0.01578611 0.18769586     589379
5 0.04763554       294549            663169 0.04243146 0.08655621     947858

As we saw in previous functions, we can also filter several rows or conditions at once:

```{r}
#| eval: FALSE
Data |> filter(Condition %in% c("PPD", "SEB")) |> head(5)
```
      bid timepoint Condition       Date infant_sex  ptype    root singletsFSC singletsSSC singletsSSCB      CD45 NotMonocytes nonDebris
1 INF0052         0       PPD 2025-07-26       Male HEU-hi 2363512     2136616     1875394      1732620 0.5873838    0.8619837 0.8429685
2 INF0100         0       PPD 2025-07-26       Male HEU-lo 2049112     1821676     1717636      1597085 0.9063081    0.9251961 0.8771889
3 INF0100         4       PPD 2025-07-26       Male HEU-lo 1063496      946587      796056       767297 0.9709891    0.9848719 0.9556049
4 INF0100         9       PPD 2025-07-26       Male HEU-lo  788368      714198      626387       600011 0.9822803    0.9842139 0.8123041
5 INF0179         0       PPD 2025-07-26       Male     HU 1380336     1242311     1047081      1000877 0.9470275    0.9575685 0.9134438
  lymphocytes      live      Dump+     Dump-    Tcells        Vd2+      Vd2-     Va7.2+    Va7.2-      CD4+      CD4-      CD8+
1   0.6408044 0.9009254 0.20743228 0.6934931 0.2835676 0.007408209 0.9925918 0.01507057 0.9775212 0.6340345 0.3434867 0.2744119
2   0.2174284 0.8929673 0.06181426 0.8311531 0.6735798 0.007137230 0.9928628 0.01671801 0.9761448 0.6145707 0.3615741 0.3312279
3   0.7313503 0.8782307 0.20727202 0.6709587 0.5989873 0.005254643 0.9947454 0.01609790 0.9786475 0.6559480 0.3226994 0.2912084
4   0.6223228 0.9566639 0.23164587 0.7250180 0.6489405 0.011935922 0.9880641 0.01855298 0.9695111 0.4306889 0.5388222 0.4908558
5   0.6996502 0.8856898 0.33186111 0.5538287 0.4441538 0.004382972 0.9956170 0.01297237 0.9826447 0.7499194 0.2327253 0.1850897
        CD8- Tcells_count lymphocytes_count  Monocytes     Debris CD45_count
1 0.06907479       184930            652155 0.13801632 0.15703150    1017713
2 0.03034620       211987            314717 0.07480391 0.12281107    1447451
3 0.03149109       326378            544883 0.01512811 0.04439511     745037
4 0.04796636       238021            366784 0.01578611 0.18769586     589379
5 0.04763554       294549            663169 0.04243146 0.08655621     947858

Or we can set up the vector first and then filter the vector:

```{r}
#| eval: FALSE
TheseConditions <- c("PPD", "SEB")
Data |> filter(Condition %in% TheseConditions) |> head(5)
```
      bid timepoint Condition       Date infant_sex  ptype    root singletsFSC singletsSSC singletsSSCB      CD45 NotMonocytes nonDebris
1 INF0052         0       PPD 2025-07-26       Male HEU-hi 2363512     2136616     1875394      1732620 0.5873838    0.8619837 0.8429685
2 INF0100         0       PPD 2025-07-26       Male HEU-lo 2049112     1821676     1717636      1597085 0.9063081    0.9251961 0.8771889
3 INF0100         4       PPD 2025-07-26       Male HEU-lo 1063496      946587      796056       767297 0.9709891    0.9848719 0.9556049
4 INF0100         9       PPD 2025-07-26       Male HEU-lo  788368      714198      626387       600011 0.9822803    0.9842139 0.8123041
5 INF0179         0       PPD 2025-07-26       Male     HU 1380336     1242311     1047081      1000877 0.9470275    0.9575685 0.9134438
  lymphocytes      live      Dump+     Dump-    Tcells        Vd2+      Vd2-     Va7.2+    Va7.2-      CD4+      CD4-      CD8+
1   0.6408044 0.9009254 0.20743228 0.6934931 0.2835676 0.007408209 0.9925918 0.01507057 0.9775212 0.6340345 0.3434867 0.2744119
2   0.2174284 0.8929673 0.06181426 0.8311531 0.6735798 0.007137230 0.9928628 0.01671801 0.9761448 0.6145707 0.3615741 0.3312279
3   0.7313503 0.8782307 0.20727202 0.6709587 0.5989873 0.005254643 0.9947454 0.01609790 0.9786475 0.6559480 0.3226994 0.2912084
4   0.6223228 0.9566639 0.23164587 0.7250180 0.6489405 0.011935922 0.9880641 0.01855298 0.9695111 0.4306889 0.5388222 0.4908558
5   0.6996502 0.8856898 0.33186111 0.5538287 0.4441538 0.004382972 0.9956170 0.01297237 0.9826447 0.7499194 0.2327253 0.1850897
        CD8- Tcells_count lymphocytes_count  Monocytes     Debris CD45_count
1 0.06907479       184930            652155 0.13801632 0.15703150    1017713
2 0.03034620       211987            314717 0.07480391 0.12281107    1447451
3 0.03149109       326378            544883 0.01512811 0.04439511     745037
4 0.04796636       238021            366784 0.01578611 0.18769586     589379
5 0.04763554       294549            663169 0.04243146 0.08655621     947858

This is OK for a low number of conditions, but when we want to select everything except one condition, it is better to use the opposite. We have to remember the formula "!", this is used in logical values to get the opposite:

```{r}
#| eval: FALSE
Subset <- Data |> filter(!Condition %in% "SEB")
Subset |> pull(Condition) |> unique()
```
[1] "Ctrl" "PPD" 

we can use "!" also with the select() function:

```{r}
#| eval: FALSE
Subset <- Data |> select(!timepoint)
Subset[1:3,]
```
      bid Condition       Date infant_sex  ptype    root singletsFSC singletsSSC singletsSSCB      CD45 NotMonocytes nonDebris
1 INF0052      Ctrl 2025-07-26       Male HEU-hi 2098368     1894070     1666179      1537396 0.5952943    0.8820349 0.8627649
2 INF0100      Ctrl 2025-07-26       Male HEU-lo 2020184     1791890     1697083      1579098 0.9106762    0.9052256 0.8602660
3 INF0100      Ctrl 2025-07-26       Male HEU-lo 1155040     1033320      875465       845446 0.9705765    0.9845400 0.9578793
  lymphocytes      live      Dump+     Dump-    Tcells        Vd2+      Vd2-     Va7.2+    Va7.2-      CD4+      CD4-      CD8+
1   0.6420138 0.9020581 0.21090996 0.6911482 0.2804264 0.008120361 0.9918796 0.01448070 0.9773989 0.6341164 0.3432825 0.2734826
2   0.2145848 0.8908981 0.06252775 0.8283703 0.6748298 0.007265620 0.9927344 0.01577499 0.9769594 0.6119112 0.3650482 0.3357696
3   0.7403110 0.8757665 0.20023803 0.6755285 0.6119129 0.004651313 0.9953487 0.01579402 0.9795547 0.6639621 0.3155925 0.2862104
        CD8- Tcells_count lymphocytes_count  Monocytes     Debris CD45_count
1 0.06979990       164771            587573 0.11796509 0.13723513     915203
2 0.02927858       208241            308583 0.09477437 0.13973396    1438047
3 0.02938209       371723            607477 0.01545999 0.04212072     820570

We can also use the mutate() function if we want to alter the content of a column or add new ones. Let’s start by transforming our current Data in to a smaller workable data.frame, highlighting the functions and pipes we learned about today:

```{r}
#| eval: FALSE
TidyData <- Data |> filter(Condition %in% "Ctrl") |> filter(timepoint %in% "0") |> select(bid, timepoint, Condition, Date, Tcells_count, CD45_count) |> rename(specimen=bid, condition=Condition) |> relocate(Date, .after=specimen)
TidyData
```
specimen       Date timepoint condition Tcells_count CD45_count
1   INF0052 2025-07-26         0      Ctrl       164771     915203
2   INF0100 2025-07-26         0      Ctrl       208241    1438047
3   INF0179 2025-07-26         0      Ctrl       291777     940733
4   INF0134 2025-07-29         0      Ctrl       127866     689676
5   INF0148 2025-07-29         0      Ctrl       234335    1013985
6   INF0191 2025-07-29         0      Ctrl        55780     715443
7   INF0124 2025-07-31         0      Ctrl        70297     687720
8   INF0149 2025-07-31         0      Ctrl       107900     857845
9   INF0169 2025-07-31         0      Ctrl        75540     854594
10  INF0019 2025-08-05         0      Ctrl       208055     873622
11  INF0032 2025-08-05         0      Ctrl       361034     753064
12  INF0180 2025-08-05         0      Ctrl       284958    1049663
13  INF0155 2025-08-07         0      Ctrl       281626    1065048
14  INF0158 2025-08-07         0      Ctrl       280913    1249338
15  INF0159 2025-08-07         0      Ctrl       452551    1190219
16  INF0013 2025-08-22         0      Ctrl       182751     836573
17  INF0023 2025-08-22         0      Ctrl       218435     968035
18  INF0030 2025-08-22         0      Ctrl        85521     732321
19  INF0166 2025-08-28         0      Ctrl       225650     739495
20  INF0199 2025-08-28         0      Ctrl       169736    1112176
21  INF0207 2025-08-28         0      Ctrl        39055     905365
22  INF0614 2025-08-30         0      Ctrl       224396    1569007
23  INF0622 2025-08-30         0      Ctrl       161924     939307

And now, we can use the mutate() function. In this example, we add a new column based on the Tcells_count and CD45_count:
```{r}
#| eval: FALSE
TidyData <- TidyData |> mutate(Tcells_ProportionCD45 = Tcells_count / CD45_count)
TidyData
```
   specimen       Date timepoint condition Tcells_count CD45_count Tcells_ProportionCD45
1   INF0052 2025-07-26         0      Ctrl       164771     915203            0.18003765
2   INF0100 2025-07-26         0      Ctrl       208241    1438047            0.14480820
3   INF0179 2025-07-26         0      Ctrl       291777     940733            0.31015921
4   INF0134 2025-07-29         0      Ctrl       127866     689676            0.18540010
5   INF0148 2025-07-29         0      Ctrl       234335    1013985            0.23110302
6   INF0191 2025-07-29         0      Ctrl        55780     715443            0.07796568
7   INF0124 2025-07-31         0      Ctrl        70297     687720            0.10221747
8   INF0149 2025-07-31         0      Ctrl       107900     857845            0.12578030
9   INF0169 2025-07-31         0      Ctrl        75540     854594            0.08839285
10  INF0019 2025-08-05         0      Ctrl       208055     873622            0.23815220
11  INF0032 2025-08-05         0      Ctrl       361034     753064            0.47942008
12  INF0180 2025-08-05         0      Ctrl       284958    1049663            0.27147570
13  INF0155 2025-08-07         0      Ctrl       281626    1065048            0.26442564
14  INF0158 2025-08-07         0      Ctrl       280913    1249338            0.22484948
15  INF0159 2025-08-07         0      Ctrl       452551    1190219            0.38022498
16  INF0013 2025-08-22         0      Ctrl       182751     836573            0.21845195
17  INF0023 2025-08-22         0      Ctrl       218435     968035            0.22564783
18  INF0030 2025-08-22         0      Ctrl        85521     732321            0.11678076
19  INF0166 2025-08-28         0      Ctrl       225650     739495            0.30514067
20  INF0199 2025-08-28         0      Ctrl       169736    1112176            0.15261613
21  INF0207 2025-08-28         0      Ctrl        39055     905365            0.04313730
22  INF0614 2025-08-30         0      Ctrl       224396    1569007            0.14301785
23  INF0622 2025-08-30         0      Ctrl       161924     939307            0.17238666

We can see that we have many significant digits being returned. Let’s round this new column to 2 significant digits by applying the round() function

```{r}
#| eval: FALSE
TidyData <- TidyData |> mutate(TcellsRounded = round(Tcells_ProportionCD45, 2))
TidyData 
```
   specimen       Date timepoint condition Tcells_count CD45_count Tcells_ProportionCD45 TcellsRounded
1   INF0052 2025-07-26         0      Ctrl       164771     915203            0.18003765          0.18
2   INF0100 2025-07-26         0      Ctrl       208241    1438047            0.14480820          0.14
3   INF0179 2025-07-26         0      Ctrl       291777     940733            0.31015921          0.31
4   INF0134 2025-07-29         0      Ctrl       127866     689676            0.18540010          0.19
5   INF0148 2025-07-29         0      Ctrl       234335    1013985            0.23110302          0.23
6   INF0191 2025-07-29         0      Ctrl        55780     715443            0.07796568          0.08
7   INF0124 2025-07-31         0      Ctrl        70297     687720            0.10221747          0.10
8   INF0149 2025-07-31         0      Ctrl       107900     857845            0.12578030          0.13
9   INF0169 2025-07-31         0      Ctrl        75540     854594            0.08839285          0.09
10  INF0019 2025-08-05         0      Ctrl       208055     873622            0.23815220          0.24
11  INF0032 2025-08-05         0      Ctrl       361034     753064            0.47942008          0.48
12  INF0180 2025-08-05         0      Ctrl       284958    1049663            0.27147570          0.27
13  INF0155 2025-08-07         0      Ctrl       281626    1065048            0.26442564          0.26
14  INF0158 2025-08-07         0      Ctrl       280913    1249338            0.22484948          0.22
15  INF0159 2025-08-07         0      Ctrl       452551    1190219            0.38022498          0.38
16  INF0013 2025-08-22         0      Ctrl       182751     836573            0.21845195          0.22
17  INF0023 2025-08-22         0      Ctrl       218435     968035            0.22564783          0.23
18  INF0030 2025-08-22         0      Ctrl        85521     732321            0.11678076          0.12
19  INF0166 2025-08-28         0      Ctrl       225650     739495            0.30514067          0.31
20  INF0199 2025-08-28         0      Ctrl       169736    1112176            0.15261613          0.15
21  INF0207 2025-08-28         0      Ctrl        39055     905365            0.04313730          0.04
22  INF0614 2025-08-30         0      Ctrl       224396    1569007            0.14301785          0.14
23  INF0622 2025-08-30         0      Ctrl       161924     939307            0.17238666          0.17

Let’s rearrange the rows so that they are descending based on the Tcell proportion. We can use this by using the desc() and arrange() functions from dplyr:

```{r}
#| eval: FALSE
TidyData <- TidyData |> arrange(desc(TcellsRounded))
TidyData
```
   specimen       Date timepoint condition Tcells_count CD45_count Tcells_ProportionCD45 TcellsRounded
1   INF0032 2025-08-05         0      Ctrl       361034     753064            0.47942008          0.48
2   INF0159 2025-08-07         0      Ctrl       452551    1190219            0.38022498          0.38
3   INF0179 2025-07-26         0      Ctrl       291777     940733            0.31015921          0.31
4   INF0166 2025-08-28         0      Ctrl       225650     739495            0.30514067          0.31
5   INF0180 2025-08-05         0      Ctrl       284958    1049663            0.27147570          0.27
6   INF0155 2025-08-07         0      Ctrl       281626    1065048            0.26442564          0.26
7   INF0019 2025-08-05         0      Ctrl       208055     873622            0.23815220          0.24
8   INF0148 2025-07-29         0      Ctrl       234335    1013985            0.23110302          0.23
9   INF0023 2025-08-22         0      Ctrl       218435     968035            0.22564783          0.23
10  INF0158 2025-08-07         0      Ctrl       280913    1249338            0.22484948          0.22
11  INF0013 2025-08-22         0      Ctrl       182751     836573            0.21845195          0.22
12  INF0134 2025-07-29         0      Ctrl       127866     689676            0.18540010          0.19
13  INF0052 2025-07-26         0      Ctrl       164771     915203            0.18003765          0.18
14  INF0622 2025-08-30         0      Ctrl       161924     939307            0.17238666          0.17
15  INF0199 2025-08-28         0      Ctrl       169736    1112176            0.15261613          0.15
16  INF0100 2025-07-26         0      Ctrl       208241    1438047            0.14480820          0.14
17  INF0614 2025-08-30         0      Ctrl       224396    1569007            0.14301785          0.14
18  INF0149 2025-07-31         0      Ctrl       107900     857845            0.12578030          0.13
19  INF0030 2025-08-22         0      Ctrl        85521     732321            0.11678076          0.12
20  INF0124 2025-07-31         0      Ctrl        70297     687720            0.10221747          0.10
21  INF0169 2025-07-31         0      Ctrl        75540     854594            0.08839285          0.09
22  INF0191 2025-07-29         0      Ctrl        55780     715443            0.07796568          0.08
23  INF0207 2025-08-28         0      Ctrl        39055     905365            0.04313730          0.04

And let’s go ahead and filter() and identify the specimens that had more than 30% T cells as part of the overall CD45 gate (context, these samples were Cord Blood):
```{r}
#| eval: FALSE
TidyData |> filter(TcellsRounded > 0.3)
```
  specimen       Date timepoint condition Tcells_count CD45_count Tcells_ProportionCD45 TcellsRounded
1  INF0032 2025-08-05         0      Ctrl       361034     753064             0.4794201          0.48
2  INF0159 2025-08-07         0      Ctrl       452551    1190219             0.3802250          0.38
3  INF0179 2025-07-26         0      Ctrl       291777     940733             0.3101592          0.31
4  INF0166 2025-08-28         0      Ctrl       225650     739495             0.3051407          0.31

If we want just the ID (specimen) from this last list of items, we can use the pull() function:

```{r}
#| eval: FALSE
TidyData |> filter(TcellsRounded > 0.3) |> pull(specimen)
```
[1] "INF0032" "INF0159" "INF0179" "INF0166"

And finally, since I may want to send the data to a supervisor, let’s go ahead and export this “tidyed” version of our data.frame out to it’s own .csv file. Working within our project folder, this would look like this:

```{r}
#| eval: FALSE
NewName <- paste0("MyNewDataset", ".csv")
StorageLocation <- file.path("C:/Users/Macarena/CytometryInR/course/week04/04_IntroToTidyverse/data", NewName)
StorageLocation
```
[1] "C:/Users/Macarena/CytometryInR/course/week04/04_IntroToTidyverse/data/MyNewDataset.csv"


```{r}
#| eval: FALSE
write.csv(TidyData, StorageLocation, row.names=FALSE)
```
