---
title: "Week8-Ways to gate"
author: "Macarena López-Mayorga"
date: 2026-06-11
format: pdf
---
First, we have to check if we have all the necessary packages installed and the proper versions:

```{r}
packageVersion("ggplot2")
```
[1] ‘4.0.3’

```{r}
packageVersion("ggcyto")
```
[1] ‘1.39.4’

```{r}
packageVersion("openCyto")
```
[1] ‘2.22.0’

```{r}
packageVersion("flowGate")
```
[1] ‘1.10.0’

This version of flowGate seems to have problems with shiny app because there are some bugs not resolved. David Rach and his colleagues have developed a forked version that fix the bug problems. So, I am going to remove flowGate package and install the version of David Rach and others:

```{r}
remove.packages("flowGate")
```
 Now, I have to call the remotes packages to install from the github repository

```{r}
library(remotes)
```

And install the forked version:
```{r}
install_github("DavidRach/flowGate")
```
Let's check now if we have the correct version:

```{r}
packageVersion("flowGate")
```
[1] ‘0.99.2’

Yes! that is correct :) . Now, we have to load in several packages that will be necessary:

```{r}
library(flowWorkspace)
library(ggplot2)
library(ggcyto)
library(flowGate)
library(openCyto)
```

```{r}
StorageLocation <- file.path("C:/Users/Macarena/CytometryInR/course/08_WaysToGate/data")
fcs_files <- list.files(StorageLocation, pattern=".fcs", full.names=TRUE)
fcs_files
```
[1] "C:/Users/Macarena/CytometryInR/course/08_WaysToGate/data/2025_07_26_AB_02_INF052_00_Ctrl.fcs"
[2] "C:/Users/Macarena/CytometryInR/course/08_WaysToGate/data/2025_07_26_AB_02_INF052_00_SEB.fcs" 
[3] "C:/Users/Macarena/CytometryInR/course/08_WaysToGate/data/2025_07_26_AB_02_INF100_00_Ctrl.fcs"
[4] "C:/Users/Macarena/CytometryInR/course/08_WaysToGate/data/2025_07_26_AB_02_INF100_00_SEB.fcs" 
[5] "C:/Users/Macarena/CytometryInR/course/08_WaysToGate/data/2025_07_26_AB_02_INF179_00_Ctrl.fcs"
[6] "C:/Users/Macarena/CytometryInR/course/08_WaysToGate/data/2025_07_26_AB_02_INF179_00_SEB.fcs" 
