---
title: "Week 05_Gating sets"
author: "Macarena López-Mayorga"
date: 2026-04-23
format: pdf
---

```{r}
Folder <- file.path("C:/Users/Macarena/CytometryInR/course/05_GatingSets/data")
fcs_files <- list.files(Folder, pattern=".fcs", full.names=TRUE)
fcs_files
```
[1] "C:/Users/Macarena/CytometryInR/course/05_GatingSets/data/2025_07_26_AB_02_INF052_00_Ctrl.fcs"
[2] "C:/Users/Macarena/CytometryInR/course/05_GatingSets/data/2025_07_26_AB_02_INF052_00_SEB.fcs" 
[3] "C:/Users/Macarena/CytometryInR/course/05_GatingSets/data/2025_07_26_AB_02_INF100_00_Ctrl.fcs"
[4] "C:/Users/Macarena/CytometryInR/course/05_GatingSets/data/2025_07_26_AB_02_INF100_00_SEB.fcs" 
[5] "C:/Users/Macarena/CytometryInR/course/05_GatingSets/data/2025_07_26_AB_02_INF179_00_Ctrl.fcs"
[6] "C:/Users/Macarena/CytometryInR/course/05_GatingSets/data/2025_07_26_AB_02_INF179_00_SEB.fcs" 

We are going to name one of them as "fcs-files[x]". In our case, the first file
```{r}
fcs_files[1]
```
[1] "C:/Users/Macarena/CytometryInR/course/05_GatingSets/data/2025_07_26_AB_02_INF052_00_Ctrl.fcs"

```{r}
library(flowCore)
```

```{r}
flowFrame <- read.FCS(filename=fcs_files[1], truncate_max_range = FALSE, transformation = FALSE)
flowFrame
```
flowFrame object '2025_07_26_AB_02_INF052_00_Ctrl.fcs'
with 10000 cells and 43 observables:
               name      desc     range  minRange    maxRange
$P1            Time        NA    896745         0     878.809
$P2           SSC-W        NA   4194304         0 4194303.000
$P3           SSC-H        NA   4194304         0 4194303.000
$P4           SSC-A        NA   4194304         0 4194303.000
$P5           FSC-W        NA   4194304         0 4194303.000
...             ...       ...       ...       ...         ...
$P39     APC-R700-A    CD107a   4194304      -111     4192506
$P40   Zombie NIR-A Viability   4194304      -111     4192506
$P41 APC-Fire 750-A      CD27   4194304      -111     4192506
$P42 APC-Fire 810-A      CCR7   4194304      -111     4192506
$P43           AF-A        NA   4194304      -111     4194303
472 keywords are stored in the 'description' slot

```{r}
str(fcs_files)
```
 chr [1:6] "C:/Users/Macarena/CytometryInR/course/05_GatingSets/data/2025_07_26_AB_02_INF052_00_Ctrl.fcs" ...

 All of them are character class.

 If we want to read multiple fcs files at once, we will need another function:
 
 ```{r}
flowSet <- read.flowSet(files=fcs_files, truncate_max_range = FALSE, transformation = FALSE)
flowSet
 ```
 A flowSet with 6 experiments.

column names(43): Time SSC-W ... APC-Fire 810-A AF-A

Also, we can select those fcs files that we can specifically read:

```{r}
read.flowSet(files=fcs_files[c(1, 3:4)], truncate_max_range = FALSE, transformation = FALSE)
```
A flowSet with 3 experiments.

column names(43): Time SSC-W ... APC-Fire 810-A AF-A

we can see that read.flowSet() has created a “flowSet” class object.

```{r}
class(flowSet)
```
[1] "flowSet"
attr(,"package")
[1] "flowCore"

In the right secondary side bar, we can see that "flowSet” is another Bioconductor style S4-type object that within its frame slot contains individual “flowFrames”.
If instead of class() we had used str(), we would have seen a similar output ton what we see in the Variables panel.

```{r}
str(flowSet)
```
Formal class 'flowSet' [package "flowCore"] with 2 slots
  ..@ frames   :<environment: 0x000001fb5d4cf5a8> 
  ..@ phenoData:Formal class 'AnnotatedDataFrame' [package "Biobase"] with 4 slots
  .. .. ..@ varMetadata      :'data.frame':	1 obs. of  1 variable:
  .. .. .. ..$ labelDescription: chr "Name"
  .. .. ..@ data             :'data.frame':	6 obs. of  1 variable:
  .. .. .. ..$ name: 'AsIs' chr [1:6] "2025_07_26_AB_02_INF052_00_Ctrl.fcs" "2025_07_26_AB_02_INF052_00_SEB.fcs" "2025_07_26_AB_02_INF100_00_Ctrl.fcs" "2025_07_26_AB_02_INF100_00_SEB.fcs" ...
  .. .. ..@ dimLabels        : chr [1:2] "rowNames" "columnNames"
  .. .. ..@ .__classVersion__:Formal class 'Versions' [package "Biobase"] with 1 slot
  .. .. .. .. ..@ .Data:List of 1
  .. .. .. .. .. ..$ : int [1:3] 1 1 0
  .. .. .. .. ..$ names: chr "AnnotatedDataFrame"

  REMINDER: While not today’s focus, remember we could access individual components inside the flowSet using the @ accessors covered during Week3.

  One consideration of this method is the contents of your .fcs files are read into your computer’s random access memory (RAM). While for individual .fcs files or small experiments this will not present a problem for most modern computers, when working with large spectral flow cytometry files containing millions of events (or trying to analyze many .fcs files at once), you may encounter situations where you can quickly exceed your computers available RAM.
  To build some contextual understanding of the problem, let’s learn how to check how much memory is being used by our individual variables/objects within our R session. We will primarily use the lobstr R packages obj_size() function, as it better handles evaluating complicated objects than base R’s object.size() function.
  
  ```{r}
object.size(flowFrame)
  ```
  3570976 bytes

  ```{r}
# Only if it is NOT installed #
install.packages("lobstr")
  ```

  ```{r}
 library(lobstr)
obj_size(flowFrame) 
  ```
  3.53 MB
  FlowSet object contains multiples flowFrame objects, so the size should be bigger:
  
  ```{r}
obj_size(flowSet)
  ```
  20.99 MB

  If we were curious how much memory total we are using within R at the current moment, we can check using the mem_used() function:
  
  ```{r}
mem_used()
  ```
  141.95 MB
  
Ultimately, how many .fcs files you are able to read in and interact with before running out of available RAM memory space will be dictated by your individual computers hardware configuration. There are various ways you can check programmatically how much RAM your computer has available, although the specific functions will vary depending on your computers operating system, since they often involve system-level code outside R. Using the ps R package’s ps_system_memory() function is one of the easier ways for Windows users.

To simplify the process, here is an additional example of where a conditional can prove useful, allowing us to check in an operating system specific manner. It takes the output of the Sys.info() function, namely the “sysname” argument and then retrieves the relavent function.

```{r}
OperatingSystem <- Sys.info()[["sysname"]]
if (OperatingSystem == "Windows")
Memory <- ps::ps_system_memory()
message("Total GB ", round(Memory$total / 1024^3, 2))
message("Free GB ", round(Memory$free / 1024^3, 2))

```
Total GB 15.78
Free GB 7.69

```{r}
install.packages("ps")
```

```{r}
library(ps)
Memory <- ps::ps_system_memory()
message("Total GB ", round(Memory$total / 1024^3, 2))
```
Total GB 15.78


```{r}
message("Free GB ", round(Memory$free / 1024^3, 2))
```
Free GB 7.83

In addition to the flowCore R package, additional flow cytometry infrastructure support is provided by the flowWorkspace package. Instead of the reading all the .fcs files contents into active RAM, flowWorkspace reduces the memory overhead by using “pointers” to interact with the object in it’s current storage location (either on your harddrive, SSD, etc.), only reading in components to RAM as needed.

```{r}
library(flowWorkspace)
```
As part of improvements to flowWorkspace, some behavior of
GatingSet objects has changed. For details, please read the section
titled "The cytoframe and cytoset classes" in the package vignette:

  vignette("flowWorkspace-Introduction", "flowWorkspace")

  ```{r}
cytoframe <- load_cytoframe_from_fcs(fcs_files[1], truncate_max_range = FALSE, transformation = FALSE)
cytoframe
  ```
cytoframe object '2025_07_26_AB_02_INF052_00_Ctrl.fcs'
with 10000 cells and 43 observables:
               name      desc       range  minRange    maxRange
$P1            Time        NA     878.809         0     878.809
$P2           SSC-W        NA 4194303.000         0 4194303.000
$P3           SSC-H        NA 4194303.000         0 4194303.000
$P4           SSC-A        NA 4194303.000         0 4194303.000
$P5           FSC-W        NA 4194303.000         0 4194303.000
...             ...       ...         ...       ...         ...
$P39     APC-R700-A    CD107a     4192506      -111     4192506
$P40   Zombie NIR-A Viability     4192506      -111     4192506
$P41 APC-Fire 750-A      CD27     4192506      -111     4192506
$P42 APC-Fire 810-A      CCR7     4192506      -111     4192506
$P43           AF-A        NA     4194303      -111     4194303
472 keywords are stored in the 'description' slot
row names(0):

```{r}
class(cytoframe)
```
[1] "cytoframe"
attr(,"package")
[1] "flowWorkspace"

Which also still errors out when not given a scalar object

```{r}
load_cytoframe_from_fcs(fcs_files, truncate_max_range = FALSE, transformation = FALSE)
```

Error:
! Expected string vector of length 1
Hide Traceback
    ▆
 1. └─flowWorkspace::load_cytoframe_from_fcs(...)
 2.   └─flowWorkspace:::parseFCS(...)

 We can use Cytoset to handle multiples fcs files:
 
 ```{r}
cytoset <- load_cytoset_from_fcs(fcs_files, truncate_max_range = FALSE, transformation = FALSE)
cytoset
 ```
 A cytoset with 6 samples.

  column names:
    Time, SSC-W, SSC-H, SSC-A, FSC-W, FSC-H, FSC-A, SSC-B-W, SSC-B-H, SSC-B-A, BUV395-A, BUV563-A, BUV615-A, BUV661-A, BUV737-A, BUV805-A, Pacific Blue-A, BV480-A, BV570-A, BV605-A, BV650-A, BV711-A, BV750-A, BV786-A, Alexa Fluor 488-A, Spark Blue 550-A, Spark Blue 574-A, RB613-A, RB705-A, RB780-A, PE-A, PE-Dazzle594-A, PE-Cy5-A, PE-Fire 700-A, PE-Fire 744-A, PE-Vio770-A, APC-A, Alexa Fluor 647-A, APC-R700-A, Zombie NIR-A, APC-Fire 750-A, APC-Fire 810-A, AF-A

```{r}
class(cytoset)
```
[1] "cytoset"
attr(,"package")
[1] "flowWorkspace"

Unlike “flowFrame” and “flowSet”, when we run str(), for “cytoframe” and “cytoset” objects we don’t get back quite as much information.
```{r}
str(cytoframe)
```
Formal class 'cytoframe' [package "flowWorkspace"] with 5 slots
  ..@ pointer    :<externalptr> 
  ..@ use.exprs  : logi TRUE
  ..@ exprs      : num[0 , 0 ] 
  ..@ parameters :Formal class 'AnnotatedDataFrame' [package "Biobase"] with 4 slots
  .. .. ..@ varMetadata      :'data.frame':	5 obs. of  1 variable:
  .. .. .. ..$ labelDescription: chr [1:5] "Name of Parameter" "Description of Parameter" "Range of Parameter" "Minimum Parameter Value after Transformation" ...
  .. .. ..@ data             :'data.frame':	0 obs. of  5 variables:
  .. .. .. ..$ name       : chr(0) 
  .. .. .. ..$ description: chr(0) 
  .. .. .. ..$ range      : num(0) 
  .. .. .. ..$ minRange   : num(0) 
  .. .. .. ..$ maxRange   : num(0) 
  .. .. ..@ dimLabels        : chr [1:2] "rowNames" "columnNames"
  .. .. ..@ .__classVersion__:Formal class 'Versions' [package "Biobase"] with 1 slot
  .. .. .. .. ..@ .Data:List of 1
  .. .. .. .. .. ..$ : int [1:3] 1 1 0
  .. .. .. .. ..$ names: chr "AnnotatedDataFrame"
  ..@ description:List of 1
  .. ..$ note: chr "empty"

  
  ```{r}
  str(cytoset)
  ```
  Formal class 'cytoset' [package "flowWorkspace"] with 3 slots
  ..@ pointer  :<externalptr> 
  ..@ frames   :<environment: 0x000001e3cf33dee8> 
  ..@ phenoData:Formal class 'AnnotatedDataFrame' [package "Biobase"] with 4 slots
  .. .. ..@ varMetadata      :'data.frame':	0 obs. of  1 variable:
  .. .. .. ..$ labelDescription: chr(0) 
  .. .. ..@ data             :'data.frame':	0 obs. of  0 variables
  .. .. ..@ dimLabels        : chr [1:2] "rowNames" "columnNames"
  .. .. ..@ .__classVersion__:Formal class 'Versions' [package "Biobase"] with 1 slot
  .. .. .. .. ..@ .Data:List of 1
  .. .. .. .. .. ..$ : int [1:3] 1 1 0
  .. .. .. .. ..$ names: chr "AnnotatedDataFrame"

  This is similarly the case when glancing at the right secondary side bar, as the respective objects under variables appear to have empty matrices where normally we would have seen the MFI values.
  Due to flowWorkspace use of pointers, the missing data remains stored on the drive, only being retrieved right before it is required. This reduces the overall RAM utilization. Let’s double check the differences in memory utilization for flowFrame/cytoframe:
  
  ```{r}
  obj_size(flowFrame)
  ```
  3.53 MB

```{r}
obj_size(cytoframe)
```
5.40 kB

And similarly the case for flowSet and cytoset:

```{r}
obj_size(flowSet)
```
20.99 MB

```{r}
obj_size(cytoset)
```
3.88 kB

Despite both R packages having been around for a while, many Bioconductor and GitHub often only implement methods to handle either flowFrames or cytoframes (although newer R packages are now allowing for both). Consequently, as we move forward in the course, it helps to be aware of which ones we are working with, and have the ability to interconvert between them as needed.

To go from a flowFrame to a cytoframe, we can use the flowFrame_to_cytoframe() function

```{r}
ConvertedToCytoframe <- flowFrame_to_cytoframe(flowFrame)
ConvertedToCytoframe
```
cytoframe object 'file207836ed2889'
with 10000 cells and 43 observables:
               name      desc       range  minRange    maxRange
$P1            Time        NA     878.809         0     878.809
$P2           SSC-W        NA 4194303.000         0 4194303.000
$P3           SSC-H        NA 4194303.000         0 4194303.000
$P4           SSC-A        NA 4194303.000         0 4194303.000
$P5           FSC-W        NA 4194303.000         0 4194303.000
...             ...       ...         ...       ...         ...
$P39     APC-R700-A    CD107a     4192506      -111     4192506
$P40   Zombie NIR-A Viability     4192506      -111     4192506
$P41 APC-Fire 750-A      CD27     4192506      -111     4192506
$P42 APC-Fire 810-A      CCR7     4192506      -111     4192506
$P43           AF-A        NA     4194303      -111     4194303
472 keywords are stored in the 'description' slot
row names(0):

As we move forward in the course, it helps to be aware of which ones we are working with, and have the ability to interconvert between them as needed.
To go from a flowFrame to a cytoframe, we can use the flowFrame_to_cytoframe() function

```{r}

```